---
title: "Microarray Analysis"
author: "Nathan Stephens"
output: html_notebook
---

DNA microarrays are used to measure the expression levels of large numbers of genes simultaneously. These experimental data come from [yeast genome microarrays](https://www.pnas.org/content/pnas/97/7/3364.full.pdf) and can be downloaded [here](http://genome-www.stanford.edu/swisnf/). In this analysis we use a two stage mixed model approach as described in [here](https://pdfs.semanticscholar.org/608a/4dc9f2464942030cb860a84ddcb215691188.pdf?_ga=2.38984291.1957266298.1552698540-1237907384.1552698540). In the first stage, a mixed model normalizes the data. In the second stage, we build independent models for each gene. **We compute approximately 7,000 models in all, one for each gene**.

### Setup

```{r setup, message = TRUE, warning = TRUE}
library(tidyverse)
library(lme4)
library(emmeans)
```

***

## Data

### Download

Download and organize the data (12 files). 

```{r}
files <- paste0("http://genome-www.stanford.edu/swisnf/",
                c("snf2ypda.txt", "snf2ypdc.txt", "snf2ypdd.txt",
                  "snf2mina.txt", "snf2minc.txt", "snf2mind.txt",
                  "swi1ypda.txt", "swi1ypdc.txt", "swi1ypdd.txt",
                  "swi1mina.txt", "swi1minc.txt", "swi1mind.txt")
                )

rawdat <- files %>%
  map(read_delim, delim = "\t") %>%
  map(rowid_to_column, var = "spot") %>%
  bind_rows(.id = "array") %>%
  filter(FLAG == 0) %>%
  mutate(name = ifelse(is.na(NAME), TYPE, NAME),
         gene = ifelse(is.na(GENE), NAME, GENE),
         type = TYPE,
         spot = as.character(spot))

treatment <- rawdat %>%
  mutate(diff = CH1I - CH1B,
         strain = case_when(array <=3 ~ "snf2rich",
                            array <=6 ~ "snf2mini",
                            array <=9 ~ "swi1rich",
                            array <=12 ~ "swi1mini"))

control <- rawdat %>%
  mutate(diff = CH2I - CH2B,
         strain = "wildtype")
 
moddat <- bind_rows(treatment, control) %>%
  filter(diff > 0) %>%
  mutate(logi = log2(diff)) %>%
  select(array, gene, name, spot, strain, logi) %>%
  arrange(gene, array, spot, strain)
```


### Normalize

Normalize the data. This takes approximately 10 seconds on Colorado.

```{r}
m1 <- lmer(logi ~ strain + (1|array/strain), moddat)

normdat <- moddat %>%
  mutate(resid = resid(m1)) %>%
  filter(!gene %in% c("EMPTY", "NORF", NA))
```

***

## Models

The following models are run sequentially.

### Gene Models

There are 7000 genes. These models take approximately 6 minutes to run in series on Colorado.

```{r}
g <- normdat %>% distinct(gene) %>% pull
out <- list()
for (i in g) {
  d <- filter(normdat, gene == i)
  k <- tryCatch({
    print(i)
    out[[i]] <- lmer(resid ~ strain + (1|spot:array), d) %>%
      emmeans(pairwise ~ strain) %>%
      .$contrasts %>%
      as_tibble() %>%
      mutate(contrast = as.character(contrast))
  }, error=function(e) cat("ERROR :",conditionMessage(e), "\n"))
}
```

### Results

Identify the significant treatment and gene interactions.

```{r}
pvals <- bind_rows(out, .id = "gene") %>%
  drop_na %>%
  mutate(n=n()) %>%
  filter(p.value < 0.05 / n) %>%
  arrange(z.ratio)
pvals
```

```{r}
ggplot(pvals, aes(p.value)) + 
  geom_density() +
  ggtitle("Distribution of signifiant p-values")
```

***

## Jobs

The following models are run using jobs.

### Gene Models (with Jobs)

```{r}
jobs <- 3
g <- normdat %>% distinct(gene) %>% pull
inds <- split(seq_along(g), cut(seq_along(g), jobs))
envs <- paste0("u", 1:jobs)
for(i in 1:jobs){
  ind <- inds[[i]]
  rstudioapi::jobRunScript("testmicro.R", 
                           workingDir = "/home/nathan/", 
                           importEnv = TRUE, 
                           exportEnv = envs[i])
}
```

### Results (from jobs)
```{r}
pvals <- mget(paste0(envs)) %>%
  map("out") %>%
  map(bind_rows) %>%
  bind_rows %>%
  drop_na %>%
  mutate(n=n()) %>%
  filter(p.value < 0.05 / n) %>%
  arrange(z.ratio)
pvals
```

```{r}
ggplot(pvals, aes(p.value)) + 
  geom_density() +
  ggtitle("Distribution of signifiant p-values")
```

### Job script ("testmicro.R")

```
library(tidyverse)
library(lme4)
library(emmeans)

out <- list()
for (i in g[ind]) {
  d <- filter(normdat, gene == i)
  k <- tryCatch({
    print(i)
    out[[i]] <- lmer(resid ~ strain + (1|spot:array), d) %>%
      emmeans(pairwise ~ strain) %>%
      .$contrasts %>%
      as_tibble() %>%
      mutate(contrast = as.character(contrast))
  }, error=function(e) cat("ERROR :",conditionMessage(e), "\n"))
}
```

***

## References

* [Download data](http://genome-www.stanford.edu/swisnf/)
* [Data explanation](https://www.pnas.org/content/pnas/97/7/3364.full.pdf)
* [Wolfinger et al](https://pdfs.semanticscholar.org/608a/4dc9f2464942030cb860a84ddcb215691188.pdf?_ga=2.38984291.1957266298.1552698540-1237907384.1552698540)
* [Y Liu](http://dl.uncw.edu/Etd/2004/liuy/yuanliu.pdf)
* [N Stephens](https://scholarsarchive.byu.edu/cgi/viewcontent.cgi?article=2114&context=etd)

